A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-07-24 manuscript/manuscript.Rmd
update manuscript

---
title             : "A tutorial on Bayesian and Frequentist Event History Analysis for psychological time-to-event data"
shorttitle        : "Tutorial on hazard modeling"

@@

abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. We discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias. The person-trial-bin oriented dataset has one row for each time bin of each trial that is at risk for event occurrence. The variable "event" in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant.

@@

@@

# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
<!-- We used `r cite_r("r-references.bib")` for all our analyses. -->

# Results

@@

>>>>

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"

@@

abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias. The person-trial-bin oriented dataset has one row for each time bin of each trial that is at risk for event occurrence. The variable "event" in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.
When creating the plots using our function plot_eha(), some warning messages will likely be generated, like these:

* Removed 2 rows containing missing values or values outside the scale range (`geom_line()`). 
* Removed 2 rows containing missing values or values outside the scale range (`geom_point()`). 
* Removed 2 rows containing missing values or values outside the scale range (`geom_segment()`).

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant.

@@

In this second tutorial we illustrate how to fit a Bayesian hazard regression model for the data set used in the first tutorial.

model formula 
analytic decisions: specification of the effect of TIME,

assumptions:
- linearity: continuous variables (TIME, trial number) 
- additivity:
- proportionality:

@@

In this third tutorial we illustrate how to fit a frequentist hazard regression model for the data set used in the first tutorial.

# Tutorial 4: Calculating descriptive statistics when there are two independent variables

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016.

@@

## Individual differences

- role of response deadlines, low-level vs. higher-level processes, 
- clustering algorithms based on h(t) and ca(t) data
-

## Cognitive psychophysiology and computational model selection

## Power analysis

- example repo on github

## Preregistration

- example preregistration for knot data

# Conclusions

### 2024-07-24 manuscript/manuscript.tex
update manuscript

\linenumbers
\usepackage{csquotes}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A tutorial on Bayesian and Frequentist Event History Analysis for psychological time-to-event data},
  pdfauthor={Sven Panis1 \& Richard Ramsey1},
  pdflang={en-EN},
  pdfkeywords={response times, event history analysis, Bayesian regression models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A tutorial on Bayesian and Frequentist Event History Analysis for psychological time-to-event data}
\author{Sven Panis\textsuperscript{1} \& Richard Ramsey\textsuperscript{1}}
\date{}

\shorttitle{Tutorial on hazard modeling}

@@

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. We discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.
}

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias. The person-trial-bin oriented dataset has one row for each time bin of each trial that is at risk for event occurrence. The variable ``event'' in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant.

@@

\caption{\label{tab:life-table}This is the life table for condition blank of participant 6.}

@@

\begin{tablenotes}[para]
\normalsize{\textit{Note.} The column named ``bin'' indicates the endpoint of each time bin (in ms), and includes time point zero. se = standard error. ca = conditional accuracy.}
\end{tablenotes}

@@

\caption{Discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of the passage of discrete waiting time.}\label{fig:eha-plot}
\end{figure}

@@

@@

\section{Methods}\label{methods}

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.

\subsection{Participants}\label{participants}

\subsection{Material}\label{material}

\subsection{Procedure}\label{procedure}

\subsection{Data analysis}\label{data-analysis}

\section{Results}\label{results}

@@

>>>>

\linenumbers
\usepackage{csquotes}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data},
  pdfauthor={Sven Panis1 \& Richard Ramsey1},
  pdflang={en-EN},
  pdfkeywords={response times, event history analysis, Bayesian regression models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data}
\author{Sven Panis\textsuperscript{1} \& Richard Ramsey\textsuperscript{1}}
\date{}

\shorttitle{Tutorial on hazard analysis}

@@

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.
}

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias. The person-trial-bin oriented dataset has one row for each time bin of each trial that is at risk for event occurrence. The variable ``event'' in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.
When plotting the data, some warning messages will be generated, like these:

\begin{itemize}
\tightlist
\item
  Removed 2 rows containing missing values or values outside the scale range (\texttt{geom\_line()}).
\item
  Removed 2 rows containing missing values or values outside the scale range (\texttt{geom\_point()}).
\item
  Removed 2 rows containing missing values or values outside the scale range (\texttt{geom\_segment()}).
\end{itemize}

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant.

@@

\caption{\label{tab:life-table}The life table for the blank prime condition of participant 6.}

@@

\begin{tablenotes}[para]
\normalsize{\textit{Note.} The column named ``bin'' indicates the endpoint of each time bin (in ms), and includes time point zero. For example the first bin is (0,40{]} with the starting point excluded and the endpoint included. se = standard error. ca = conditional accuracy.}
\end{tablenotes}

@@

\caption{Estimated discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of the passage of discrete waiting time.}\label{fig:eha-plot}
\end{figure}

@@

In this second tutorial we illustrate how to fit a Bayesian hazard regression model for the data set used in the first tutorial.

@@

In this third tutorial we illustrate how to fit a frequentist hazard regression model for the data set used in the first tutorial.

\section{Tutorial 4: Calculating descriptive statistics when there are two independent variables}\label{tutorial-4-calculating-descriptive-statistics-when-there-are-two-independent-variables}

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016).

@@

\subsection{Individual differences}\label{individual-differences}

\begin{itemize}
\tightlist
\item
  role of response deadlines, low-level vs.~higher-level processes,
\item
  clustering algorithms based on h(t) and ca(t) data
\item
\end{itemize}

\subsection{Cognitive psychophysiology and computational model selection}\label{cognitive-psychophysiology-and-computational-model-selection}

\subsection{Power analysis}\label{power-analysis}

\begin{itemize}
\tightlist
\item
  example repo on github
\end{itemize}

\subsection{Preregistration}\label{preregistration}

\begin{itemize}
\tightlist
\item
  example preregistration for knot data
\end{itemize}

\section{Conclusions}\label{conclusions}
